---
title: "R Sourse Codes: Analysis of the Stationarity and Correlation of the Global Temperature and Carbon Dioxide Time Series"
author: "Upul Rupassara"
date: "2023-04-19"
output:
  pdf_document: default
  html_document: default
---


### NOTE: WHEN RUNING THE CODES, REMOVE THE "#" IN FRONT OF "INSTALL.PACKAGES()". 


# Install Packages
```{r}
#install.packages("ggplot2")
library(ggplot2)
```

```{r}
#install.packages("astsa")
library(astsa)
```


###Transforming data into time series data

```{r}


#install.packages("xts")
library(xts)

```

### Transforming the data into timeseries format

```{r}

tc = read.csv("tc.csv")
x = tc$temperature
y = tc$co2
z = tc$ca

temp = ts(x, frequency = 1, start = 1960, end = 2022) 
co2 = ts(y, frequency = 1, start = 1960, end = 2022)
ca = ts(z, frequency = 1, start = 1960, end = 2022)


par(mfrow = c(3,1))
par(mar=c(2, 2, 2, 2), mfrow=c(2,1),oma = c(1, 1, 1, 1))
par(adj = 0)

```


# Model 1:
```{r}
# Temperature
model1 = lm(temp ~ time(temp), na.action = NULL)
model1
summary(model1)

```
## AIC for Model 1:
```{r}
# Temperature
AIC(model1)/nrow(tc) - log(2*pi)
```
## BIC for Model 1
```{r}
# Temperature
BIC(model1)/nrow(tc) - log(2*pi)
```
## SSE for Model 1
```{r}
# Temperature
sum(resid(model1)^2)
```
### AIC with formula, model 1
```{r}
k = 2
log(sum(resid(model1)^2)/nrow(tc)) + (nrow(tc)+(2*k))/(nrow(tc)) 
```

```{r}
nrow(tc)
```

### MSEfor model 1 using summary of the model 1
```{r}
mean(summary(model1)$residuals^2)
```
```{r}
summary(model1)$df
```
```{r}
sum(summary(model1)$residual^2)
```
### MSE model 1
```{r}
sum(summary(model1)$residual^2)/61
```
### AICc with formula, model 1
```{r}
k = 2
log(sum(resid(model1)^2)/nrow(tc)) + (nrow(tc)+k)/(nrow(tc)-k-2) 
```

### BIC with formula, model 1

```{r}
k = 2
log(sum(resid(model1)^2)/nrow(tc)) + k*log(nrow(tc))/(nrow(tc)) 
```
# Model 2
```{r}

# Temp dependent on CO2

cc = mean(co2)
C = co2 - cc
model2 = lm(temp ~ time(temp) + C)
model2
summary(model2)
```

```{r}
sum(summary(model2)$residual^2)
```
### CALCULATION (F-STATISTICS)
```{r}
(0.632 - 0.495)/(0.495/60)
qf(0.001, 1, 60, lower.tail = FALSE)
```
### AIC with formula, model 2
```{r}
k = 3
log(sum(resid(model2)^2)/nrow(tc)) + (nrow(tc)+2*k)/(nrow(tc)) 
```
### AIC (ALTERNATIVE METHOD)
```{r}
AIC(model2)/nrow(tc) - log(2*pi)
```
### AICc with formula, model 2
```{r}
k = 3
log(sum(resid(model2)^2)/nrow(tc)) + (nrow(tc)+k)/(nrow(tc)-k-2) 
```

### BIC with formula, model 2
```{r}
k = 3
r=  (resid(model2))^2
log(sum(r)/nrow(tc)) + k*log(nrow(tc))/(nrow(tc)) 
```

```{r}
(BIC(model2))/nrow(tc) - log(2*pi)
```
### Smoothing splines for temprature
### SOURCE CODES FOR FIGURE 2

```{r}

#png("fig02.png", width = 12, height = 10, units = 'cm', res = 300)
plot(x = tc$Year, y = tc$temperature, type = "o", xlab = "Year", ylab = "Temperature Anomaly")

lines(smooth.spline(time(temp), temp, spar = 0.5), lwd = 2, col = 4)
abline(model1, col = 2, lty = "dashed", lwd = 1.75)
legend(1960, 1, legend=c("Smoothing Splines", "Trend"),
       col=c("blue", "red"), lty=1:2, cex=0.8)

```

### Durbin-Watson test (Optional)
```{r}
library(lmtest)
dwtest(formula = model1,  alternative = "two.sided")
```
### Temperature Detrended and differenced

### SOURCE CODES FOR FIGURE 3

```{r}

#png("fig03.png", width = 12, height = 12, units = 'cm', res = 300)
par(mfrow = c(2,1))
par(mar=c(2, 2, 2, 2), mfrow=c(2,1),oma = c(1, 1, 1, 1)) #margin a nd outer margin
par(adj = 0) #shift the title to left

plot(resid(model1), type = "o", main = expression(a.~Detrended~temperature~series), pch = 19, cex = 0.5, col = 4)

abline(h = 0, lty = 2, col = "gray")
plot(diff(temp), type = "o", main = expression(b.~Differenced~temperature~series), pch = 19, cex =0.5, col = 2)
abline(h = 0, lty = 2, col = "gray")
```

### ACF plots
### SOURCE CODES FOR FIGURE 4

```{r}
#png("fig04.png", width = 16, height = 12, units = 'cm', res = 300)
par(mar=c(3, 3, 3, 3), mfrow=c(3,1),oma = c(1, 1, 1, 1)) #margin a nd outer margin
par(adj = 0) #shift the title to left
acf(temp, 10, main = "a. ACF vs Lag: temperature anomaly", col = 6, lwd = 2)
acf(resid(model1),10, main = "b. ACF vs Lag: detrended temperature series", col = 2, lwd = 2)
acf(diff(temp), 10,main = "c. ACF vs Lag: differenced temperature series", col = 3, lwd = 2)
```

### CO2 time series ACF plots

```{r}
modelc = lm(co2~time(co2), na.action = NULL)
summary(modelc)
```


### Smoothing Splines for Carbon dioxide

### SOURCE CODES FOR FIGURE 5

```{r}

#png("fig05.png", width = 12, height = 10, units = 'cm', res = 300)
plot(x = tc$Year, y = tc$co2, type = "o", xlab = "Year", ylab = "Carbon Dioxide")
lines(smooth.spline(time(co2), co2, spar = 0.5), lwd = 2, col = 4)
abline(modelc, col = 2, lty = "dashed", lwd = 1.75)
legend(1960, 420, legend=c("Smoothing Splines", "Trend"), col=c("blue", "red"), lty=1:2, cex=0.8)

```

### CO2 
### SOURCE CODES FOR FIGURE 6

```{r}

#png("fig06.png", width = 12, height = 12,  units = 'cm', res = 300)
par(mar=c(2, 2, 2, 2), mfrow=c(3,1),oma = c(1, 1, 1, 1))
par(adj = 0)

modelc = lm(co2~time(co2), na.action = NULL)


#mat_layout<- matrix(c(1,0,2,0), ncol = 2)
#layout(mat_layout)

plot(resid(modelc), main = "a. CO2 detrended series", type = "o", pch = 19, col = 4)


#grid(nx = NULL, ny = NULL,lty = 2, col = "gray", lwd = 2)
abline(h = 0, lty = 2, col = "gray")
plot(diff(co2), main = "b. CO2 differenced series (d = 1)", type = "o",  pch = 19, col = 2)
abline(h = 0, lty = 2, col = "gray")
plot(diff(diff(co2)), type = "o", main = "c. CO2 differenced series (d = 2)", pch = 19, col = 3)
abline(h = 0, lty = 2, col = "gray")

```



### SOURCE CODES FOR FIGURE 7

```{r}
#png("fig07.png", width = 16, height = 12, units = 'cm', res = 300)

par(mar=c(3, 3, 3, 3), mfrow=c(3,1),oma = c(1, 1, 1, 1)) #margin a nd outer margin
par(adj = 0) #shift the title to left
acf(tc$co2, 10, main = "a. ACF vs Lag: CO2 series", col = "6", lwd = 2)
acf(resid(modelc),10, main = "b. ACF vs Lag: CO2 detrended series", col = 2, lwd = 2 )
acf(diff(tc$co2), 10, main = "c. ACF vs Lag: CO2 differenced series", col = 2, lwd = 2)

```

### REFERENCE FOR ADJUSTING GAP###
https://stackoverflow.com/questions/15848942/how-to-reduce-space-gap-between-multiple-graphs-in-r


### DW test for CO2 (OPTIONAL)
```{r}
dwtest(formula = modelc,  alternative = "two.sided")
```

### DW test for Temperature (OPTIONAL)
```{r}
dwtest(formula = model1,  alternative = "two.sided")
```
According to the ACF plots detrended series represents the stationary behavior than the differenced series.

### Lag plot

`DW test

H0: residuals are independent
H1: Residuals are not independent

```{r}
dwtest(formula = model1, alternative = "two.sided")
```

```{r}
dwtest(model2, alternative = "two.sided")
```

According to the Durbin-Watson test the second model follows the assumptions than th first model. The assumption of residuals are independent.
### ACF for CO2 second differenced series

### SOURCE CODES FOR FIGURE 8


```{r}
#png("fig08.png", width = 12, height = 8, units = 'cm', res = 300)
#par(mfrow = c(1,2))
#acf(diff(tc$co2), 10, main = "c. ACF vs Lag: CO2 differenced series", col = 2, lwd = 2)
acf(diff(diff(tc$co2)), 10, main = " ", col = 2, lwd = 2, xlab = " ")
```
### SOURCE CODES FOR FIGURE 9

```{r}
#png("fig09.png", width = 12, height = 10, units = 'cm', res = 300)
lag= tc$co2
lag1.plot(lag, 4, col = 4)
mean(tc$co2)


```


### Lag Regression

We need the following package to align the time series before making the model.

```{r}
#install.packages("dynlm")
library(dynlm)
```



```{r}
dwtest(model1, alternative = "two.sided")
```

### SOURCE CODES FOR FIGURE 10
```{r}
#png("fig10.png", width = 12, height = 10, units = 'cm', res = 300)
par(mfrow = c(2, 2))
plot(model2, col = 4)
```

### Cross  correlation

### SOURCE CODES FOR FIGURE 11

```{r}
#png("fig11.png", width = 14, height = 8, units = 'cm', res = 300)
x = co2
y = lag(x,-1) + temp
ccf(y,x, ylab = "Cross-correlation", type = "correlation", col = 4, lwd = 2, main = " ")
grid(NA, 5, lwd = 2) # grid only in y-direction
abline(v = 0, col = 2, lty = 2, lwd = 3)
legend(-15, 1, expression(Temperature~leads), cex=0.6)
legend(10, 1, expression(CO2~leads), cex=0.6)
```

```{r}
bgtest(model1, order = 1)
```


```{r}
#install.packages("tseries")
library(tseries)
```
### Dicky Fuller test (OPTIONAL)

### REFERENCE
https://atsa-es.github.io/atsa-labs/sec-boxjenkins-aug-dickey-fuller.html


```{r}
#install.packages("urca")
library(urca)
```

```{r}
library(urca)
s= ur.df(temp, type = "trend")
s
summary(s)
```

